---
title: "Practical Machine Learning - Course project"

---

###Synopsis
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

###Summary
A random forest model was build with an accuracy of 0.994, so the out of sample error is 0.006. When run against the testing set, the model predicted all 20 cases correctly.

###Data Processing
####Load data
```{r setup, cache=TRUE, echo=TRUE, results='hide'}
if(!file.exists("./data")){dir.create("./data")}
# download and import training data
fileUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileUrl,destfile="./data/training.csv",method="curl")

# import the first five rows to get an idea of the data
tab5rows<-read.table("./data/training.csv", quote='"', sep=',', header=TRUE, nrows = 5)
dim(tab5rows)
str(tab5rows)

# get the full data set
basicTrainingData<-read.csv("./data/training.csv",sep=",",header=TRUE)


# download and import testing data
fileUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl,destfile="./data/testing.csv",method="curl")
basicTestingData<-read.csv("./data/testing.csv",sep=",",header=TRUE)

# explore variables
dim(basicTrainingData)
str(basicTrainingData)
```

####Clean data
```{r cleaning, echo=TRUE, results='hide'}
# remove the first seven rows which contain no predictive values
cutTrainingData<-basicTrainingData[,-(1:7)]

# delete columns with all missing values
noNAsTrainingData<-cutTrainingData[,colSums(is.na(cutTrainingData)) == 0]

# exclude near zero value predictors
library(caret)
nsv_positions<-nearZeroVar(noNAsTrainingData,saveMetrics=FALSE)
# filter out the variables that were near-zero-value.
nsvTrainingData<-noNAsTrainingData[-c(nsv_positions)]

# find correlated variables to exclude one of the highly correlated pairs
filtered_training_no_class<-nsvTrainingData[-c(dim(nsvTrainingData))]
correlated_cols_to_exclude<-findCorrelation(cor(filtered_training_no_class), cutoff= 0.8)
finalTrainingData<-nsvTrainingData[-c(correlated_cols_to_exclude)]

#explore the clean dataset
dim(finalTrainingData)
str(finalTrainingData)

# transform the test set accordingly
colNames<-names(finalTrainingData)
finalTestingData<-basicTestingData[,names(basicTestingData) %in% c(colNames)]

```

####Build random forests
```{r model, echo=TRUE, results='hide'}
#split the training set
part<-createDataPartition(finalTrainingData$classe, p=0.7, list=F)
training<-finalTrainingData[part,]
probe<-finalTrainingData[-part,]
str(training)

set.seed(1023)
#train a random forest model
library(randomForest)
rf.model<-randomForest(classe~.,data=training,importance=TRUE)
```

####Validate the model
```{r validate, echo=TRUE, results='markup'}
yhat.rf<-predict(rf.model,newdata=probe)
confusionMatrix(yhat.rf,probe$classe)
```

####Test the model on the testing data set
```{r testing, echo=TRUE, results='markup'}
ytest.rf<-predict(rf.model,newdata=finalTestingData)
```

####Generate files for submission
```{r submit,echo=TRUE,results='hide'}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(ytest.rf)
```


